---
title: "Simulacrum"
author: "Timothy Nguyen"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, results = 'hide'}
# consider using renv for library management
library(dplyr)
library(dtplyr) # data.table wrapper for dplyr for faster execution
library(knitr)
library(ggplot2)
library(tidyr)

```

# Data ingestion

```{r}
ZIP_FILEPATH <- "./simulacrum_v2.1.0.zip"
zip_tables <- unzip(ZIP_FILEPATH, list = TRUE)

patient <- read.csv(unz(ZIP_FILEPATH, "simulacrum_v2.1.0/Data/sim_av_patient.csv"))
tumour  <- read.csv(unz(ZIP_FILEPATH, "simulacrum_v2.1.0/Data/sim_av_tumour.csv"))

```

# Missingness checks

**Note**: Missing values are empty string for character data type, else `NA` for numeric

```{r}
check_missing_var <- function(x) {
  if (is.character(x)) return(sum(x == ''))
  else return(sum(is.na(x)))
}

check_missing_tbl <- function(.data) {
  # Return summary table with count and percent of missing values
  missing <- lapply(
    .data,
    check_missing_var
  )
  
  missing_tbl <- data.frame(
    variable = names(missing),
    type = unname(sapply(.data, typeof)),
    n_missing = as.numeric(missing),
    pct_missing = as.numeric(missing) / nrow(.data)
  )
  
  return(missing_tbl)
}

format_missing_tbl <- function(.data) {
  # Same output as `check_missing_tbl` but formatted slightly for presentation
  return(
    check_missing_tbl(.data) |>
      arrange(desc(n_missing)) |>
      mutate(n_missing = scales::comma(n_missing),
             pct_missing = scales::percent(pct_missing, accuracy = .1)) |>
      rename(`Missing (count)` = n_missing,
             `Missing (percent)` = pct_missing)
  )
}
missing_patient <- check_missing_tbl(patient)
missing_tumour  <- check_missing_tbl(tumour)
```

_Patient table_

* Death code variables have the highest missingness. This reflects the majority of patients recorded as being alive. For patients that died, most records have the underlying cause of death (2.1% missing) and location of death recorded (1.8% missing).

* Ethnicity is missing in 4.8% of patient records followed by date in which vital status is recorded (0.1%).

* There is no missingness for patient ID, patient gender, vital status, and link number (proxy for NHS number).

```{r}
kable(format_missing_tbl(patient))

# Additional analysis: missingness for patients who have died
kable(format_missing_tbl(patient |> filter(grepl("^D", VITALSTATUS))),
      caption = "Filtered for patients who have died")
```

_Tumour table_

Missingness is high for some variables where data is only recorded if it is relevant to the tumour. For example, the Gleason Tertiary score (`GLEASON_TERTIARY`) has the highest missingness (99.7%) however this is only relevant for prostate cancers. This is similarly the case for progesterone receptor status (`PR_STATUS`, 94% missing) and oestrogen receptor status (`ER_STATUS`, 91.2% missing).

```{r}
kable(format_missing_tbl(tumour))
```

# Feature engineering

```{r}
map_gender_label <- function(x) {
  case_when(
    {{ x }} == "1" ~ "Male",
    {{ x }} == "2" ~ "Female",
    {{ x }} == "9" ~ "Indeterminate",
    TRUE ~ NA_character_
    )
}

patient_transform <- patient |>
  dtplyr::lazy_dt() |>
  mutate(VITALSTATUSDATE = as.Date(VITALSTATUSDATE),
         VITALSTATUSYEAR = lubridate::year(VITALSTATUSDATE),
         # map to labels for downstream analysis
         # see all_z_lookup_tables.xlsx
         GENDER_LABEL = map_gender_label(GENDER)) |>
  as_tibble()

patient_transform |>
  select(VITALSTATUSDATE, VITALSTATUSYEAR) |>
  head()

# expect no missingness in gender
stopifnot(sum(is.na(patient_transform$GENDER_LABEL)) == 0)
```

# Analysis

## Age and sex distribution

**Note**: Age is recorded at time of diagnosis. The same patient may appear multiple times if they have more than one tumour. Tumour diagnosis dates from 1 January 2016 to 31 December 2019, inclusive.

Male patients are typically slightly older than female patients at the time of tumour diagnosis (Male median: 71 years; Female median: 68 years). Only 2 patients with indeterminate sex were recorded in the tumour dataset.

```{r}
tumour_transform <- tumour |>
  mutate(GENDER_LABEL = map_gender_label(GENDER),
         DIAGNOSISDATEBEST = as.Date(DIAGNOSISDATEBEST))

table(tumour_transform$GENDER_LABEL)
tapply(tumour_transform$AGE, tumour_transform$GENDER_LABEL, summary)
```

```{r}
# check date range of analysis
MIN_DATE <- min(tumour_transform$DIAGNOSISDATEBEST, na.rm = TRUE)
MAX_DATE <- max(tumour_transform$DIAGNOSISDATEBEST, na.rm = TRUE)
```

```{r}
# Exclude indeterminate sex because small cell sizes (n = 2)
ggplot(data = tumour_transform |> filter(GENDER_LABEL != "Indeterminate"),
       aes(x = AGE)) +
  geom_histogram(binwidth = 5) +
  facet_grid(. ~ GENDER_LABEL) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Distribution of patient and sex",
       subtitle = "Age as of tumour diagnosis date",
       x = "Age", y = "Count")
```

## Comorbidities

* Most tumour diagnosed patients (78.3%) do not have any recorded chronic condition (i.e. total Charlson comorbidity score of zero).

* The most common comorbidities contributing to the Charlson score were pulmonary disease, diabetes, and renal disease.

```{r}
charlson_total <- tumour_transform |>
  count(CHRL_TOT_27_03) |>
  mutate(pct = n / sum(n),
         # omit percent labels for small values
         pct_label = if_else(pct < 0.001, NA_real_, pct)) 

ggplot(data = charlson_total, aes(x = CHRL_TOT_27_03, y = pct,
                                  label = scales::percent(pct_label, accuracy = .1))) +
  geom_col() +
  geom_text(vjust = -0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = min(charlson_total$CHRL_TOT_27_03, na.rm = TRUE):
                       max(charlson_total$CHRL_TOT_27_03, na.rm = TRUE)) +
  labs(x = "Total Charlson comorbidity score", y = "Percent",
       title = "Distribution of total Charlson comorbidity score",
       subtitle = "Tumour diagnosed patients from 1 Jan 2016 to 31 Dec 2019")
```

```{r}
# Separate comma separated comorbidity flags and obtain underlying comorbidities
# contributing to total Charlson comorbidity score

# determine how many columns to separate into
max_flags <- tumour_transform |>
  select(COMORBIDITIES_27_03) |>
  mutate(len = stringr::str_count(COMORBIDITIES_27_03, ",")) |>
  summarise(max_len = max(len)) |>
  pull()

wide_cols <- paste0("comorbidity_", 1:(max_flags + 1))
charlson_breakdown <- tumour_transform |>
  select(COMORBIDITIES_27_03) |>
  tidyr::separate_wider_delim(
    cols = COMORBIDITIES_27_03,
    delim = ",",
    names = wide_cols,
    too_few = "align_start"
  )

comorbidity_map <- readxl::read_xlsx("Documents/all_z_lookup_tables.xlsx",
                                     sheet = "z_comorbidities") |>
  select(Code, Description)

charlson_long <- charlson_breakdown |>
  tidyr::pivot_longer(cols = all_of(wide_cols), values_drop_na = TRUE) |>
  filter(value != "") |>
  mutate(value = as.double(value)) |>
  left_join(comorbidity_map, by = c("value" = "Code"))

```

```{r}
charlson_flag_summary <- charlson_long |>
  count(Description) |>
  arrange(desc(n))

ggplot(data = charlson_flag_summary, aes(x = reorder(Description, n), y = n,
                                         label = paste0(round(n / 1e3), "k"))) +
  geom_col() +
  geom_text(hjust = 1) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = '', y = 'Count', title = 'Count of Charlson comorbidity flags',
       subtitle = "Tumour diagnosed patients from 1 Jan 2016 to 31 Dec 2019")
```

## Ethnicity for lowest income

```{r}
# Get ethnic groups for lowest quintile income level group
ethnicity_map <- readxl::read_xlsx("Documents/all_z_lookup_tables.xlsx",
                                     sheet = "z_ethnicity") |>
  select(Code, Description)

# seems like patient's quintiles may change at time of tumour diagnosis
check_unique_income <- tumour_transform |>
  distinct(PATIENTID, QUINTILE_2019) |>
  count(PATIENTID) |>
  filter(n > 1)
  
tumour_patient <- tumour_transform |>
  left_join(patient_transform |> select(PATIENTID, ETHNICITY), by = "PATIENTID") |>
  left_join(ethnicity_map, by = c("ETHNICITY" = "Code"))
stopifnot(nrow(tumour_patient) == nrow(tumour_transform)) # expect 1-to-1

# let's free some memory
rm(tumour)
gc()
```

```{r}
# (a) Percent of ethnic groups for most deprived income group
ethnicity_count <- tumour_patient |>
  filter(QUINTILE_2019 == "1 - most deprived") |>
  count(Description) |>
  mutate(Percent = scales::percent(n / sum(n), accuracy = .1)) |>
  arrange(desc(n))

ethnicity_count |>
  select(Ethnicity = Description,
         Count = n, 
         Percent) |>
  kable()
```

* 
```{r}
# (b) Which ethnic groups have higher % of low income earners
ethnicity_income <- tumour_patient |>
  count(QUINTILE_2019, Description) |>
  group_by(Description) |>
  mutate(Percent = n / sum(n),
         Percent_label = scales::percent(Percent, accuracy = .1))|>
  ungroup() |>
  filter(QUINTILE_2019 == "1 - most deprived") |>
  arrange(desc(Percent))

ethnicity_income |>
  select(Ethnicity = Description,
         Count = n, 
         Percent = Percent_label) |>
  kable()
```
