---
title: "Simulacrum"
author: "Timothy Nguyen"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, results = 'hide'}
library(dplyr)
library(knitr)

```

# Data ingestion

```{r}
ZIP_FILEPATH <- "./simulacrum_v2.1.0.zip"
zip_tables <- unzip(ZIP_FILEPATH, list = TRUE)

patient <- read.csv(unz(ZIP_FILEPATH, "simulacrum_v2.1.0/Data/sim_av_patient.csv"))
tumour  <- read.csv(unz(ZIP_FILEPATH, "simulacrum_v2.1.0/Data/sim_av_tumour.csv"))

```

# Missingness checks

**Note**: Missing values are empty string for character data type, else `NA` for numeric

```{r}
check_missing_var <- function(x) {
  if (is.character(x)) return(sum(x == ''))
  else return(sum(is.na(x)))
}

check_missing_tbl <- function(.data) {
  # Return summary table with count and percent of missing values
  missing <- lapply(
    .data,
    check_missing_var
  )
  
  missing_tbl <- data.frame(
    variable = names(missing),
    type = unname(sapply(.data, typeof)),
    n_missing = as.numeric(missing),
    pct_missing = as.numeric(missing) / nrow(.data)
  )
  
  return(missing_tbl)
}

format_missing_tbl <- function(.data) {
  # Same output as `check_missing_tbl` but formatted slightly for presentation
  return(
    check_missing_tbl(.data) |>
      arrange(desc(n_missing)) |>
      mutate(n_missing = scales::comma(n_missing),
             pct_missing = scales::percent(pct_missing, accuracy = .1)) |>
      rename(`Missing (count)` = n_missing,
             `Missing (percent)` = pct_missing)
  )
}
missing_patient <- check_missing_tbl(patient)
missing_tumours <- check_missing_tbl(tumours)
```

_Patient table_

* Death code variables have the highest missingness. This reflects the majority of patients recorded as being alive. For patients that died, most records have the underlying cause of death (2.1% missing) and location of death recorded (1.8% missing).

* Ethnicity is missing in 4.8% of patient records followed by date in which vital status is recorded (0.1%).

* There is no missingness for patient ID, patient gender, vital status, and link number (proxy for NHS number).

```{r}
kable(format_missing_tbl(patient))

# Additional analysis: missingness for patients who have died
kable(format_missing_tbl(patient |> filter(grepl("^D", VITALSTATUS))),
      caption = "Filtered for patients who have died")
```

_Tumour table_

Missingness is high for some variables where data is only recorded if it is relevant to the tumour. For example, the Gleason Tertiary score (`GLEASON_TERTIARY`) has the highest missingness (99.7%) however this is only relevant for prostate cancers. This is similarly the case for progesterone receptor status (`PR_STATUS`, 94% missing) and oestrogen receptor status (`ER_STATUS`, 91.2% missing).

```{r}
kable(format_missing_tbl(tumour))
```

